---
title: "Reviewer Expertise, Engagement, and Peer Review of Grants"
subtitle: "Working Session with CIHR's Funding Analytics Team"
author: "Sam Harper & Arijit Nandi"
institute: "McGill University"
date: last-modified
date-format: iso
format:
  revealjs:
    theme: [default, custom.scss]
    width: 1400
    height: 788
    slide-number: true
    html-math-method: mathjax
    # embed-resources: true
editor: source
---

```{r setup, include=FALSE}
library(here)
library(tidyverse)
library(tinytable)
knitr::opts_chunk$set(echo = FALSE)
```

## Why we're here

- RoR grant funded: **The Influence of Reviewer Expertise and Engagement on Peer Review of Grants**
- Before we finalize a data-sharing and analysis-plan agreement, we want to work through this together:
  - what we're proposing to estimate, and why
  - what we've already tested on simulated data
  - where we anticipate friction, technically and logistically

::: aside
Application 1017-2025-00465. Funded Sept 2025 -- Aug 2028.
:::

## The research question

Overall aim: does the CIHR Project Grant peer review process affect **reliability and fairness** in scoring and funding success?

1. **Aim 1** -- How do reviewer expertise and engagement affect the impact of panel discussion on scores?
2. **Aim 2** -- Do these effects differ by applicant gender or career stage?
3. **Aim 3** (exploratory) -- How would alternative funding-decision schemes (e.g. weighting by engagement, partial randomization in the grey zone) compare to the status quo?

## {.center}

### [How does a score become a score?]{.blue}

## The process we're modeling

- 3 assigned reviewers read the application, score it, discuss, and agree a **consensus score**
- All panel members -- including the 3 reviewers themselves -- then give a **final score**
- Final score must be within **+/- 0.5** of consensus, entered to **one decimal place**
- The average of final scores (equal weight, all members) feeds the funding decision

::: aside
Confirming our understanding is correct here is one of today's questions.
:::

## Our primary outcome

$$d_{ijk} = \text{final score} - \text{consensus score}$$

for the $i$th panel member, $j$th application, $k$th committee

::: incremental
- Bounded by design: $d_{ijk} \in [-0.5, 0.5]$
- A member who simply adopts consensus has $d_{ijk} = 0$ -- this is most members, most of the time
- We model the *change*, not the final score level, because:
  - it matches CIHR's actual +/- 0.5 rule exactly (an implicit slope of 1, not an estimated one)
  - it keeps "no change" at a fixed value (0) across every application, regardless of where that application's consensus happened to land
:::

## {.center}

### [Testing the design before we ask you to run anything]{.blue}

## Why we built a simulation first

- Code has to run correctly **remotely**, without interactive access to real data
- So: simulate data with the same structure CIHR would provide, build and pressure-test the full pipeline against it, *then* hand over code we're confident in
- Mirrors 50 committees x 15 discussed applications x 24 members

## Simulated data structure

```{r sim-data-table}
d <- read_csv(here("data", "sim-deviation-data.csv"), show_col_types = FALSE)

d |>
  select(cmte, app, memno, job, exp, consensus, deviated, deviation, score) |>
  slice_head(n = 6) |>
  tt(caption = "First 6 rows of the simulated committee-application-member dataset")
```

- `job`: reviewer vs. non-reviewing panelist
- `exp`: self-described expertise (high / medium / low / not enough)
- `consensus`, `score`: rounded to the nearest tenth, +/- 0.5 bound enforced

## A constraint we initially missed

- CIHR scores are entered to **one decimal place only** -- our first simulation draft didn't enforce this
- Fixed it, which had a real downstream consequence: the magnitude of deviation among those who *do* deviate is not continuous -- it's one of **10 discrete steps** (+/-0.1 ... +/-0.5)

```{r deviation-table}
d |>
  filter(deviated == 1) |>
  count(deviation) |>
  mutate(deviation = sprintf("%+.1f", deviation)) |>
  pivot_wider(names_from = deviation, values_from = n) |>
  tt(caption = "Distribution of deviation magnitude among deviators (simulated data)")
```

::: aside
Sharply peaked near +/-0.1, sparse at +/-0.5 -- exactly where funding-decision consequences (Aim 3) are largest.
:::

## Why not a single linear model on $d_{ijk}$?

Checked this empirically on the simulated data before committing to anything more complex:

- A plain `lm(deviation ~ job + exp)` finds **no meaningful effect** of expertise or engagement (largest coefficient: 0.005 points, mostly non-significant)
- A `glm(deviated ~ job + exp, family = binomial)` on the same covariates recovers the **known, large, highly significant** effect built into the simulation
- Because deviation direction is symmetric around zero, almost all the covariate signal lives in *whether* someone deviates, not in the average size of the deviation -- a mean-only model structurally can't see it

## Model structure

Two linked models (brms has no single native family for "signed, bounded, spike at an interior value"):

::: {.larger}
**Part 1 -- did this member deviate at all?**
:::

$$\text{deviated}_{ijk} \sim \text{Bernoulli}(p_{ijk})$$

::: {.larger}
**Part 2 -- if they deviated, by how much?**
:::

Ordinal `cumulative()` model over the 10 discrete steps (not treated as continuous)

$$E[d_{ijk}] = P(\text{deviate}) \times E[\text{magnitude} \mid \text{deviate}]$$

## Software

- **Bayesian**, via `brms` + `cmdstanr` (Stan backend)
- Same conventions as our other multilevel work: weakly-informative priors, prior-predictive checks before fitting, `marginaleffects` for contrasts
- Random effects for committee, reviewer, and application throughout

## {.center}

### [Questions for the Funding Analytics Team]{.red}

## Can your environment run this?

- `brms`/`cmdstanr` compiles a Stan model in C++ at fit time -- this is our single biggest technical risk
- We can supply pre-compiled executables, but that raises portability questions of its own
- **Question:** does your remote execution environment have a C++ toolchain available, or would we need another path (e.g. a Docker image, a pre-approved compute environment)?

## Which fields can you actually extract?

Fields we're assuming are available, still to be confirmed:

- Initial reviewer scores, consensus score, final panel-member scores
- Self-described expertise, engagement (reviewer/panelist), gender, experience, career stage, past funding success, conflicts of interest
- Applicant gender, career stage
- Keywords/domain, re-submission status, PRC, number of investigators, funding requested

**Question:** which of these are extractable at the panel-member level (not just application level), and linked correctly across the consensus -> final score transition?

## Policy changes we need your help accounting for

Digging into CIHR's public documentation on the Project Grant program surfaced several structural changes that any historical/trend data will need to account for:

::: {.smaller-text}
| Year | Change |
|---|---|
| 2016 | Equalization introduced (early-career researcher NPIs) |
| 2019 | Foundation Grant program sunset -- gradual return of holders into Project Grant pool, still ongoing |
| 2021 | Equalization expanded (female NPIs, French-language); ~$50M/competition added |
| Fall 2022 | Some peer review committees dissolved/merged |
| 2022 | Indigenous Health Research given a separate iterative review process -- excluded from general ranking/equalization |
| Fall 2024 | Equalization expanded (racialized, disability); equalization zone redefined (~15-34% -> 10-25%) |
| Mar 2026 | Active consultation on further changes (in-person format, committee mandates, single annual competition) -- undecided |
:::

## Policy changes (continued)

**Questions:**

- For any historical/trend data: how should we treat the pre/post-2019 applicant pool shift as Foundation holders returned?
- Should Indigenous Health Research committee data be excluded from our models entirely, given it isn't part of the general ranking system?
- Is there a stable committee identifier we can use across the Fall 2022 restructuring?
- Given the March 2026 consultation is still open, is there a chance committee structure or review format changes again during our data-collection window (2025--2028)?

## Logistics

- Confidentiality workflow: we supply code, your team runs it and returns results -- no direct access to identifiers
- **Question:** how should we budget your analyst time for the remote runs, including any back-and-forth on debugging?
- **Question:** what's a realistic timeline for finalizing the data-sharing and analysis-plan agreement?
- We'd like to propose a first Ottawa visit early in Year 1 to work through the details of confidential fields and the analysis plan directly

## {.center}

### [Discussion]{.blue}

Thank you -- looking forward to working through this together.

sam.harper@mcgill.ca
