---
title: "Simulation Design and Modeling Strategy"
subtitle: "The Influence of Reviewer Expertise and Engagement on Peer Review of Grants"
author: "Sam Harper & Arijit Nandi"
date: last-modified
date-format: iso
format:
  html:
    toc: true
    code-fold: true
    code-tools: true
    embed-resources: true
    df-print: paged
    fontsize: 14pt
    mainfont: "Roboto Slab"
    include-in-header:
      text: |
        <link href="https://fonts.googleapis.com/css2?family=Roboto+Slab:wght@300;400;600&display=swap" rel="stylesheet">
        <style>
          body, p, li, td, th { font-family: 'Roboto Slab', serif !important; font-weight: 400; color: rgb(54, 55, 55); }
          h1, h2, h3, h4 { font-family: 'Roboto Slab', serif !important; font-weight: 600; color: rgb(54, 55, 55); }
          body { line-height: 1.8; }
        </style>
  typst:
    mainfont: "C059"
    fontsize: "11pt"
    font-paths:
      - fonts
      - writing/fonts
    margin:
      x: "1.87cm"
      y: "1.87cm"
    keep-typ: true
    citeproc: true
    bibliographystyle: vancouver.csl
    toc: true
    include-in-header:
      text: |
        #show heading.where(level: 1): set text(size: 1.25em)
        #show heading.where(level: 2): set text(size: 1.1em)
        #show heading.where(level: 3): set text(size: 1em)
bibliography: ror.bib
---

```{r setup, include=FALSE}
library(here)
library(tidyverse)
library(modelsummary)
library(tinytable)
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

# Purpose of this document

This is a design and validation exercise for a project that aims to assess how reviewer engagement and expertise may affect grant scores during CIHR peer review. Given the restrictive nature of CIHR funding data, this document does not use real CIHR data; rather it lays out the data-generating process we believe matches CIHR's Project Grant peer review, the model we intend to fit against it, and evidence that the models we will use can actually recover simulated effects. 

It is not a full pre-analysis plan (no pre-specified hypotheses, stopping
rules, or multiplicity strategy yet), but it's close in spirit, and a
subsequent revision could become one.

**Status and audience.** This draft is for Arijit first -- comments on the
outcome definition, the two-part model structure, the Aim 2 interaction
specification, or the priors are all welcome before this goes anywhere else.
A revised version will go to CIHR's Funding Analytics Team (Matt Hogel and
colleagues) as part of finalizing the data-sharing and analysis-plan
agreement. It supersedes the `sim-data.qmd` note shared with CIHR earlier,
which predates the move to a Bayesian framework and the current two-part,
ordinal outcome model -- the underlying data structure it simulates is very
different from what follows here.

# Research questions

Full detail is in the funded proposal (`cihr-ror-proposal.qmd`); briefly:

1. **Aim 1** -- How do reviewer expertise (self-described: high/medium/low/not
   enough) and engagement (assigned reviewer vs. non-reviewing panelist)
   affect the impact of panel discussion on scores?
2. **Aim 2** -- Do these effects differ by applicant characteristics, namely
   gender or career stage?
3. **Aim 3** (exploratory) -- How would alternative funding-decision schemes
   (e.g. reweighting by engagement, partial randomization near the funding
   threshold) compare to the status quo?

# The review process we're modeling

- The 3 assigned reviewers read the application, score it, discuss, and
  agree a **consensus score**.
- All panel members -- including the 3 reviewers themselves -- then submit a
  **final score**. Reviewers are not bound to their own consensus number;
  they can move too.
- The final score must fall within **&plusmn;0.5** of the consensus score,
  and is entered to **one decimal place**.
- The (equally weighted) average of final scores across all panel members
  feeds the funding decision.

Confirming this understanding is correct is itself one of the open items
for CIHR (see below) -- everything downstream depends on it.

# Data structure

We're assuming a three-level structure: committees, discussed applications
nested in committees, and panel members nested in committees (crossed with
applications, since a member reviews many applications within a cycle).
Based on 2024 Project Grant committee sizes, our working numbers are
**50 committees &times; 15 discussed applications per committee &times; 24
members per committee**.

CIHR's Funding Analytics Team confirmed by email (2025-12-04) which fields
are actually extractable for a data pull, versus fields that can only ever
be touched by a CIHR analyst running our code in-house on the real data
(full table and follow-up questions in `code/ror-research-log.qmd`).
Headline for this document: committee/application/member identifiers, role
(reviewer vs. panelist), self-described expertise, initial reviewer scores,
consensus score, final scores, and funding result are all extractable.
**Applicant gender and career stage -- Aim 2's entire basis -- are not**,
and won't appear even in CIHR's own distribution-matched dummy data. That
constraint is why this document exists: the simulation below is the only
rehearsal Aim 2's code gets before it runs once, unsupervised, on real data.

# Primary outcome: modeling the change, not the level

Let $d_{ijk}$ be the final score minus the consensus score, for the
$i$th panel member on the $j$th application in the $k$th committee:

$$d_{ijk} = \text{final score}_{ijk} - \text{consensus score}_{jk}$$

We model $d_{ijk}$ directly, rather than modeling the final score with
consensus score as a covariate, for three reasons:

1. **The point mass at zero doesn't disappear if you model the final score
   instead -- it just stops being a fixed reference.** A member who doesn't
   move from consensus has $d_{ijk}=0$ regardless of which application
   they're scoring; a `final_score ~ consensus_score + ...` model still has
   that spike, just relocated to wherever a given application's consensus
   happened to land. Differencing standardizes the spike's location across
   every application, which is what makes a shared two-part model
   tractable.
2. **Avoids reintroducing a confound.** A freely estimated slope on
   consensus in a `final_score ~ consensus + ...` model has to recover
   something close to 1 from the data; if reviewer composition happens to
   correlate with consensus level across committees, that estimation can
   leak into the coefficients we actually care about. Differencing with an
   implicit slope of exactly 1 -- the true administrative rule, not an
   assumption -- removes that channel by construction.
3. **Matches the actual research question.** Aim 1 is about the *change*
   induced by discussion, not the *level* of the final score (driven mostly
   by application quality, which isn't under study here). Nothing is lost:
   $\text{final score} = \text{consensus} + d_{ijk}$ is a trivial identity,
   so anything Aim 3 needs at the score level for funding-decision
   simulations can be reconstructed downstream.

# Why a two-part model

We expect $d_{ijk}$ to have a spike at exactly zero -- members who simply
adopt the consensus score -- plus continuous variation among those who
depart from it. `brms` has no single native family for "point mass at an
interior value plus continuous elsewhere" the way `hurdle_poisson()`
handles zero counts, since $d_{ijk}$ is signed and bounded
($\pm 0.5$), not non-negative. So the outcome is split into two linked
models:

1. **Did this member deviate at all?** A Bernoulli model on
   $\mathbb{1}[d_{ijk} \neq 0]$.
2. **If they deviated, by how much?** A model for the (signed) magnitude,
   conditional on deviating.

$$E[d_{ijk}] = P(\text{deviate}) \times E[\text{magnitude} \mid \text{deviate}]$$

This isn't just theoretically motivated -- we checked it empirically before
committing to the extra complexity. If most of the covariate signal about
reviewer engagement/expertise actually lived in the *average size* of the
deviation rather than *whether* a deviation happens, a single linear model
on $d_{ijk}$ would be simpler and would suffice.

```{r aim1-linear-vs-two-part}
#| tbl-cap: "Aim 1: a linear model on raw deviation vs. a logistic model on whether any deviation occurred, same covariates, same simulated data"

d1 <- read_csv(here("data", "sim-deviation-data.csv"), show_col_types = FALSE)

m_linear <- lm(deviation ~ job + exp, data = d1)
m_logistic <- glm(deviated ~ job + exp, data = d1, family = binomial())

modelsummary(
  list("Linear: deviation" = m_linear, "Logistic: deviated (0/1)" = m_logistic),
  gof_omit = "DF|Deviance|R2|AIC|BIC|RMSE|Log.Lik|F",
  escape = TRUE
)
```

The linear model's coefficients on `job`/`exp` are tiny and mostly
non-significant (largest around 0.005 points, on an outcome bounded at
$\pm 0.5$). The logistic model on the same covariates recovers the large,
highly significant effect pattern actually built into the simulator
(`code/ror-sim-deviation.R`). Because deviation magnitude is roughly
symmetric around zero conditional on deviating, essentially all the
covariate signal lives in *whether* someone deviates -- a model that only
looks at the conditional mean of $d_{ijk}$ structurally cannot see it. This
confirms the two-part structure is doing real work here, not adding
complexity for its own sake.

# Model specification (Aim 1)

## Simulating the data-generating process

Fifty committees, 15 discussed applications per committee, 24 members per
committee; 3 of the 24 members on each application are the assigned
reviewers, the rest are non-reviewing panelists. A consensus score is drawn
per application (committee- and application-level random effects only, no
member-level noise, since it's agreed before any individual scoring
happens). Whether each member deviates from that consensus is a function of
their role and self-described expertise; if they deviate, the signed
magnitude is drawn from a truncated distribution and rounded to the nearest
tenth, matching CIHR's one-decimal-place scoring (with rejection sampling
so a "deviated" row can never round down to a contradictory zero).

```{r aim1-params}
#| eval: false
#| code-summary: "code/ror-sim-deviation.R -- parameters and generative process"

# define parameters
cmte_n   = 50     # number of committees
app_n    = 15     # number of discussed applications per committee
mem_n    = 24     # number of committee members per committee

b0       = 4.1    # intercept for consensus score
u0c_sd   = 0.1    # random intercept SD for committee (consensus level)
u0a_sd   = 0.3    # random intercept SD for application (consensus level)

# probability of *any* deviation from consensus (logit scale) --
# this is the "hu"-equivalent part
a0       = -0.8   # baseline logit prob. of deviating
a1       =  0.3   # panelist vs. reviewer
a2       = -0.4   # high expertise
a3       =  0.2   # low expertise
a4       =  0.5   # no expertise

# signed magnitude of deviation, given deviation occurs, truncated to
# +/- 0.5 (CIHR's stated bound on final vs. consensus score)
dev_bias = 0      # placeholder: no systematic direction yet
dev_sd   = 0.15
dev_min  = -0.5
dev_max  =  0.5

# ... committee/application/member structure, reviewer assignment, and
# expertise assignment omitted here -- see code/ror-sim-deviation.R for
# the complete script

data <- data |>
  mutate(
    p_dev = plogis(a0 + (a1 * panelist) + (a2 * exp_high) +
      (a3 * exp_low) + (a4 * exp_none)),
    deviated = rbinom(n(), 1, p_dev),
    deviation = if_else(
      deviated == 1,
      round_tenth(rtruncnorm(n(), a = dev_min, b = dev_max,
        mean = dev_bias, sd = dev_sd)),
      0)
  )
```

Every parameter above is an illustrative placeholder, not an estimate from
real data -- there is no real data yet. The point of the simulation is
structural (does our modeling approach recover a known effect of a given
size?), not predictive of what CIHR's actual numbers will look like.

## Part 1: did this member deviate?

```{r aim1-m1}
#| eval: false
#| code-summary: "code/ror-analysis-score-models.R -- Part 1 (bernoulli)"

m1_deviate <-
  brm(data = d1,
      family = bernoulli(),
      deviated ~ 1 + job + exp + (1 | cmte) + (1 | cid) + (1 | app),
      prior = c(prior(normal(0, 1.5), class = Intercept),
                prior(normal(0, 0.5), class = b),
                prior(exponential(1), class = sd)),
      iter = 2000, warmup = 1000, chains = 4, cores = 4,
      sample_prior = "yes",
      control = list(adapt_delta = 0.95))
```

Random intercepts for committee, member (`cid`), and application; fixed
effects for role and expertise. Weakly informative priors throughout --
`normal(0, 1.5)` on the (logit-scale) intercept, `normal(0, 0.5)` on
coefficients, `exponential(1)` on group-level SDs -- matching the
conventions established in our other multilevel Bayesian work
[e.g. @nandi2020]. `sample_prior = "yes"` so we always have prior-predictive
draws to check against before trusting the posterior.

## Part 2: how large, given a deviation?

CIHR scores are entered to one decimal place, so a deviator's score can only
land on one of 10 discrete steps ($\pm0.1 \ldots \pm0.5$ relative to
consensus). That's genuinely ordinal/discrete data, not a continuous
quantity with occasional extreme values -- a continuous family
(`student()`/`gaussian()`) would assign probability to impossible values
and has no natural bound at $\pm0.5$ without truncation hacks. We use an
ordinal `cumulative()` model instead, with flexible (non-equidistant)
thresholds, since there's no reason to assume the 10 steps are equally
likely -- and in the simulated data they clearly aren't (concentrated near
$\pm0.1$, sparse at $\pm0.5$, which happen to be the swings most relevant to
Aim 3's funding-decision question).

```{r aim1-m2}
#| eval: false
#| code-summary: "code/ror-analysis-score-models.R -- Part 2 (ordinal)"

m1_magnitude <-
  brm(data = d1_dev,
      family = cumulative(link = "logit", threshold = "flexible"),
      deviation ~ 1 + job + exp + (1 | cmte) + (1 | cid) + (1 | app),
      prior = c(prior(normal(0, 1.5), class = Intercept),
                prior(normal(0, 0.5), class = b),
                prior(exponential(1), class = sd)),
      iter = 2000, warmup = 1000, chains = 4, cores = 4,
      sample_prior = "yes",
      control = list(adapt_delta = 0.95))
```

The cost: $E[\text{magnitude} \mid \text{deviate}]$ from an ordinal model is
not a linear prediction -- it's a probability-weighted sum over the 10
category values, computed per posterior draw, not read off a default
`marginaleffects` contrast. That combination step (and the corresponding
$E[d_{ijk}] = P(\text{deviate}) \times E[\text{magnitude} \mid
\text{deviate}]$ calculation across both models) is written but not yet
implemented in code -- see Open Questions below.

## Software

Bayesian throughout, via `brms` + `cmdstanr` (Stan backend), matching the
conventions in our other multilevel work. `marginaleffects` for posterior
contrasts; models cached to `code/fits/` and deliberately re-fit (not
silently reused) whenever formula or data change.

# Extending to Aim 2: applicant characteristics

Since applicant gender and career stage will never be extractable -- not
even as CIHR's own dummy data -- this simulation is the only pre-deployment
test that code touching those variables gets. We added them as
**application-level** attributes (one value per application, like
consensus score, not per member) to a new, separate simulation script
(`code/ror-sim-aim2.R`) rather than growing the Aim 1 script further --
see "A note on simulation architecture" below.

Critically, we didn't just add *main* effects of gender/career stage. Aim
2's actual question is whether the *engagement/expertise effect on
deviation* differs by applicant characteristic -- which requires an
**interaction** term to exist in the generating process, or there's nothing
for the fitted model to recover.

```{r aim2-params}
#| eval: false
#| code-summary: "code/ror-sim-aim2.R -- applicant-level additions"

# applicant-level main effects (modest -- not the effects of interest)
g1       =  0.1   # applicant gender = female
c1       = -0.1   # applicant career stage = early

# Aim 2's actual target: does the reviewer-engagement / expertise effect
# on P(deviate) differ by applicant gender / career stage? Baked in large
# on purpose so we can confirm the two-part model machinery actually
# recovers a real interaction, not just a real main effect.
i_panelist_female     =  0.5  # job(panelist) x gender(female)
i_expnone_earlycareer =  0.6  # exp(none) x career_stage(early)

data <- data |>
  mutate(
    p_dev = plogis(a0 + (a1 * panelist) + (a2 * exp_high) +
      (a3 * exp_low) + (a4 * exp_none) +
      (g1 * female) + (c1 * early_career) +
      (i_panelist_female * panelist * female) +
      (i_expnone_earlycareer * exp_none * early_career)),
    deviated = rbinom(n(), 1, p_dev)
    # ... deviation magnitude drawn the same way as Aim 1; applicant
    # terms are deliberately left out of the magnitude part, consistent
    # with the Aim 1 finding above that signal lives in whether someone
    # deviates, not in the average size of the deviation
  )
```

**To be clear about what these numbers are and aren't:** they are chosen to
be large enough to detect cleanly in a simulation of this size, not
estimates -- or even guesses -- about the true magnitude or even direction
of any real effect in CIHR review. The goal is purely to confirm our
statistical machinery would notice an interaction if the real data
contained one.

```{r aim2-interaction-check}
#| tbl-cap: "Aim 2: interaction coefficients recovered from the simulated data (n = 18,000)"

d2 <- read_csv(here("data", "sim-aim2-data.csv"), show_col_types = FALSE)

m_aim2 <- glm(deviated ~ job * gender + exp * career_stage,
  data = d2, family = binomial())

coefs <- summary(m_aim2)$coefficients
coefs[grepl(":", rownames(coefs)), ] |>
  as_tibble(rownames = "term") |>
  tt(digits = 3)
```

Both interactions are recovered cleanly and with high significance (the
sign flips relative to the parameters above are just R's alphabetical
choice of reference category -- `job`'s reference is "panelist", `gender`'s
is "female", `career_stage`'s is "early", so e.g.
`jobreviewer:gendermale` is algebraically the same interaction as
`job(panelist):gender(female)`, just from the opposite corner of the 2x2).

**Not yet built:** an Aim-2-specific version of
`code/ror-analysis-score-models.R` that actually includes these interaction
terms in the `brm()` formulas. The simulation is ready; the model script
that would be tested against it isn't written yet.

# Aim 3: not yet designed

Aim 3 asks how alternative funding-decision schemes -- reweighting scores
by engagement, or partially randomizing funding decisions for applications
near the threshold -- would compare to the status quo. That's a
structurally different kind of simulation: an intervention/counterfactual
layered on top of the funding decision itself, not just an extension of
Aims 1-2's data-generating process for reviewer behavior. We haven't
started designing it, and would welcome input on what the alternative
schemes worth simulating actually are before building anything.

# A note on simulation architecture

We're deliberately keeping a separate, self-contained simulation script per
aim (`ror-sim-deviation.R` for Aim 1, `ror-sim-aim2.R` for Aim 2, and
eventually an Aim 3 script) rather than one script that keeps growing.
Some duplication of the committee/application/member setup across scripts
is an accepted cost of each one being independently runnable and easy to
reason about in isolation.

# What this simulation validates, and what it doesn't

It validates that: (1) the two-part model structure is necessary, not
over-engineering, for this outcome; (2) the ordinal magnitude submodel is
the right match for CIHR's one-decimal-place scoring; (3) the model can
recover both main effects (Aim 1) and interaction effects (Aim 2) of a
detectable size, using only the fields CIHR has confirmed are extractable
plus the fields that would only ever be touched via an in-house CIHR-run
analysis.

It does **not** tell us anything about the true size, direction, or even
existence of any of these effects in actual CIHR peer review -- every
parameter in both simulations is an illustrative placeholder chosen for
statistical detectability, not drawn from evidence.

# Open questions

## For Arijit

- Is $d_{ijk}$ (final $-$ consensus) the right primary outcome, or is there
  a better way to frame the estimand for Aim 1/2?
- Does the interaction specification for Aim 2 (`job x gender`,
  `exp x career_stage`, both entering only the deviation-probability part)
  match how you'd operationalize the research question, or would you
  frame it differently?
- Any concerns about the priors, the random-effects structure, or the
  choice of an ordinal rather than continuous family for the magnitude
  part?
- Anything about the overall statistical strategy that looks wrong before
  this goes to CIHR?

## For CIHR's Funding Analytics Team

(Carried over from `code/ror-research-log.qmd`; repeated here since this
document is the more likely vehicle for actually sending them.)

1. For fields that can't be extracted (applicant gender/career stage,
   reviewer gender/experience, etc.) -- can the dummy-data step still
   include *synthetic* versions of those columns for us to pipeline-test
   against, or is the in-house-run path a black box until the real run?
2. Resubmission status is only available from 2023 onward -- does that
   mean our cohort has to start at the 2023 competition if we want that
   field, or can it be requested as a partially-missing covariate over a
   longer window?
3. How should we define/measure "reviewer experience" for the fields where
   that's the blocker to a yes/no on extractability?
4. Can you confirm our understanding of the review process itself (the
   $\pm0.5$ rule, one-decimal-place scoring, equal-weight averaging into
   the funding decision) is accurate?
5. How much variation exists in the number of members across committees,     applications per committee, and scores per application (given conflicts)?
6. Discussion time? Do we know how long the discussion period goes at the application level?

# What comes next

Once the Aim 2 model script exists and `FIT_MODELS` is switched on -- after
confirming CIHR's execution environment can actually compile/run Stan via
`cmdstanr`, and after prior-predictive checks on both submodels -- this
document (or its next revision) will add a results section presenting
posterior marginal effects via `marginaleffects` for both the
deviation-probability and magnitude submodels, and showing recovery of the
known simulated effects at the full posterior level, the same way the
frequentist checks above do at a point-estimate level.

# References

::: {#refs}
:::
