---
title: "Causal considerations for the RoR project"
subtitle: 
author: "Sam Harper"
date: last-modified
date-format: iso
format:
  html:
    toc: true
    embed-resources: true
    code-fold: show
execute:
  warning: false
  message: false
---

## Purpose

A working document for the causal structure underlying this project --
not a finished DAG, a place to think through it and keep the reasoning
alongside the code that draws it, so it's easy to pick back up later.
See `code/ror-research-log.qmd` (2026-09-02 entry) for the discussion
that prompted this.

```{r}
#| label: setup
library(here)
library(dagitty)
source(here("code", "dag-helpers.R"))
```

`dag-helpers.R` vendors `drawdag()` and friends from Richard
McElreath's `rethinking` package
([source](https://github.com/rmcelreath/rethinking/blob/master/R/drawdag.R),
GPL-3) rather than depending on the whole package, which pulls in
rstan/cmdstanr build machinery not needed just to plot a DAG. It's
built on `dagitty` for DAG specification/layout and `shape::Arrows()`
for the curved arrowheads.

## Aim 1: `exp` / `job` / `deviated`

`job` is a **mediator**, not a confounder, of `exp`'s effect on
`deviated` -- CIHR assigns reviewer/panelist role partly based on
self-declared expertise, and expertise may also affect deviation
behavior directly. Confirmed this structure is already baked into the
simulation itself (`exp_reviewer_weight` in `ror-sim-deviate.R`), not
just a concern about real data -- see the research log for the full
reasoning and an empirical check on `data/sim-deviate.csv`.

```{r}
#| label: dag-aim1
#| fig-cap: "exp -> job -> deviated, exp -> deviated direct"

dag_aim1 <- dagitty("dag {
  exp -> job
  exp -> deviated
  job -> deviated
}")

coordinates(dag_aim1) <- list(
  x = c(exp = 0, job = 1, deviated = 2),
  y = c(exp = 0, job = -1, deviated = 0)
)

drawdag(dag_aim1, cex = 1.3)
```

`dagitty` confirms the adjustment sets match the by-hand reasoning in
the research log:

```{r}
#| label: adjustment-sets

cat("total effect of exp on deviated:\n")
print(adjustmentSets(dag_aim1, exposure = "exp", outcome = "deviated", effect = "total"))

cat("\ndirect effect of exp on deviated (not through job):\n")
print(adjustmentSets(dag_aim1, exposure = "exp", outcome = "deviated", effect = "direct"))

cat("\ntotal effect of job on deviated:\n")
print(adjustmentSets(dag_aim1, exposure = "job", outcome = "deviated", effect = "total"))
```

- `exp`'s *total* effect needs no adjustment -- matches the plain
  `deviated ~ exp` model (`m1` in the research log check).
- `exp`'s *direct* effect (net of role) needs `{job}` -- matches
  `deviated ~ exp + job` (`m2`).
- `job`'s effect needs `{exp}` -- `exp` is a genuine confounder of the
  `job -> deviated` relationship (a common cause of both), so `m2` is
  also the right specification for `job`'s own coefficient.

## Adding applicant characteristics as precision covariates

`job` stays the focal exposure. Applicant characteristics (gender,
past funding success, age) plausibly predict the outcome but can't
plausibly affect anything about the reviewer side (`job`, `exp`) --
nor could reviewer assignment retroactively cause an applicant's
gender or age. That makes them exogenous to the whole reviewer-side
process, entering only through `deviated`: a textbook "precision
covariate" pattern, not a confounder -- a variable d-separated from
the exposure can be added to reduce residual variance without opening
any biasing path (Cinelli, Forney & Pearl's "good control" case,
opposite of the `exp`/`job` mediator situation above).

**Two things this doesn't settle, worth remembering before trusting
the precision gain:** (1) Tamblyn (2018) found applicant
characteristics predict *level* of final score, not (as far as we
know) *deviation* from consensus -- our actual outcome. Whether the
precision argument really holds for `deviated` specifically is
untested; could build a small, deliberate applicant-characteristic
effect into the simulation (same pattern as Aim 2's `job:gender`/
`exp:career_stage` interactions) and check empirically. (2) our models
are logistic/ordinal, not linear -- adding a genuine outcome predictor
in a non-collapsible link typically shifts the *point estimate's
magnitude* too, not just the SE, unlike the OLS intuition. (3) CIHR
data-extractability is unconfirmed for these three specifically --
applicant gender/career stage are already confirmed *not* extractable
(see `code/ror-research-log.qmd`, 2026-08-04); past funding success
and age haven't been checked with Matthew Hogel.

This is also conceptually distinct from Aim 2: these are *main-effect*
precision covariates for `job`'s estimand here, not the
effect-modification question Aim 2 actually asks (does `job`/`exp`'s
*effect* differ *by* applicant characteristic). Both could coexist in
the same model.

```{r}
#| label: dag-aim1-precision
#| fig-cap: "job as focal exposure, applicant characteristics as precision covariates on deviated only"

dag_aim1_precision <- dagitty("dag {
  exp -> job
  exp -> deviated
  job -> deviated
  applicant_gender -> deviated
  applicant_funding -> deviated
  applicant_age -> deviated
}")

exposures(dag_aim1_precision) <- "job"
outcomes(dag_aim1_precision)  <- "deviated"

coordinates(dag_aim1_precision) <- list(
  x = c(exp = 0, job = 1, deviated = 2,
        applicant_gender = 1.3, applicant_funding = 2, applicant_age = 2.7),
  y = c(exp = 0, job = -1.5, deviated = 0,
        applicant_gender = 1.5, applicant_funding = 2, applicant_age = 1.5)
)

drawdag(dag_aim1_precision, cex = 1.1)
```

```{r}
#| label: adjustment-sets-precision

cat("total effect of job on deviated:\n")
print(adjustmentSets(dag_aim1_precision))
```

Adjustment set for `job` is unchanged (`{exp}`) -- adding the
applicant nodes doesn't change *what* needs to be conditioned on for
identification, only (potentially) how precisely it's estimated, which
is exactly the point of a precision covariate: it shouldn't change the
adjustment set at all.

## Open questions / where to take this next

- Test the precision-covariate claim empirically once applicant
  characteristics are actually in the simulation (Aim 2's sim already
  has `gender`/`career_stage` -- could check SE(`job`) with/without
  them as main effects on `deviated`, separate from Aim 2's interaction
  question).
- Check with Matthew Hogel whether "past funding success" and "age"
  are extractable, independent of gender/career stage's already-known
  status.
-
